\input pmacro \reportformat
\centerline{\titlefont Compiling AGAMA with MS Visual Studio}

\bigskip\noindent
Here I explain how to get Vasiliev's  AGAMA package (Action-based  Galaxy
Modelling Framework) running with Microsoft's freely-distributed `Visual Studio'
compiler. This is essential if you want to run AGAMA under
Windows because it is impossible to compile either Python or the Gnu
Scientific Library with the Gnu compilers, and a single compiler must compile all
components of a code.

Instructions for installing GSL can be found at

\noindent https://solarianprogrammer.com/2020/01/26/getting-started-gsl-gnu-scientific-library-windows-macos-linux/

In a Linux/Gnu environment, AGAMA is embodied in shared library {\tt agama.so}. Under
VS it is contained in two files {\tt agama.lib} and {\tt agama.dll}. The former is
what the linker uses, while the latter is required at runtime, so has to lie
on your {\tt path}. I have the following file {\tt vscl.bat} on my {\tt path}
so I can compile and link a program {\tt testit.cpp} that uses AGAMA by
typing {\tt vscl testit}
\begintt
@echo off
cl.exe /c /openmp /I\u\c\agama\agama -EHsc %1.cpp
@if not %ERRORLEVEL% == 0 (goto end)
link %1.obj agama.lib /LIBPATH:\u\c\agama\agama\x64\release /NOIMPLIB
del %1.obj
del %1.exp
:end
\endtt
 What follows {\tt /I} in the second line is the path to my AGAMA source code
\& what follows {\tt LIBPATH:} in the fourth line is the path to {\tt
agama.lib}. That's where the VS IDE put it: I used the VS IDE to compile
AGAMA rather than hacking the file {\tt makefile.local} distributed with
AGAMA and using {\tt make}. The IDE handles access to the Gnu Scientific
Library and Python if they were installed using the {\tt vcpkg} utility.  The
last {\tt del} command is necessary because VS insists on producing an
`export' library file {\tt.exp} even when told to make an executable. 

\bigskip\centerline{\bf Installing AGAMA}

\noindent Untar everything into your source directory and start the Visual
Studio IDE and via the  File tab  open {\tt agama.sln} (sln stands for `solution') within the
source directory. Then provided you've installed the Gnu Scientific Library
within VS, you should be able to `build agama' from within the `build' tab of
the VS IDL. VS is a bossy compiler so there will be zillions of warnings. It
will even declare an error on correct code, such one using {\tt fopen} rather
than {\tt fopen\_s}.  It's easier to identify any errors if you use the
drop-down menu at the top centre of the error list to change to `Build only'.


You may need to set some compiler options. At the bottom of the Project tab, click on
Properties. Then click on C/C++ and by clicking on open MP support get Yes (/openmp)

After building AGAMA with Visual Studio Community 2019, Version 16.9.2 I
tested the procedure for installing AGAMA by transferring the code to an
older machine running Version 16.4.2. I encountered two problems: (i) min was
said not to be a member of sdt -- I fixed the problem by adding {\tt
\#include <algorithm>} to {\tt math\_base.h}; (ii) the linker complained
about twoPhase template instantiation. This problem was fixed by clicking
Project$\rightarrow$Properties$\rightarrow$C/C++$\rightarrow$Command line and
typing in {\tt /Zc:twoPhase-} at the bottom.  This click sequence could also
be used to establish what the compiler and linker flags should be if you
prefer to hack {\tt makefile.local} rather than using the IDE.
 

Among the many outputs of an error-free compilation will be {\tt agama.dll},
which you need to copy to a folder that lies on your path, and {\tt
agama.lib}. The outputs will be in {\tt x64$\backslash$release} or {\tt
x64$\backslash$debug} if you've set VS to debugging (top left centre of IDE).
Here I'm assuming that you have a 64-bit machine and that you will be running
the code from the `x64 Native Tools Command Prompt' that should be in your
Visual Studio folder on your start menu. If you have a 32-bit machine, change
the IDE from `x64' to `x86' at the top. 

The following lines in a makefile
\begintt
ASRC = \u\c\agama\agama
ALIB  = \u\c\agama\agama\x64\release\agama.lib
ADLL  = \u\c\agama\agama\x64\release\agama.dll
testit.exe: testit.cpp  $(ALIB) $(ADLL) other.obj
	cl /openmp /I$(ASRC) -EHsc testit.cpp /link /out:testit.exe NOIMPLIB \ 
	other.obj \libs\press.lib $(ALIB)
\endtt
will update testit.exe when any of {\tt testit.cpp}, {\tt other.obj} or AGAMA is changed. {\tt testit} will
be linked to AGAMA, other.obj and the static library
{\tt$\backslash$libs$\backslash$press.lib}. The words before {\tt/link} are instructions
to the compiler, while those after are instructions to the linker.

For some reason VS doesn't put any symbols from the dll it is constructing into the
{\tt.lib} file unless explicitly instructed to do so. So any symbol you want to reference
from code that's not in the AGAMA library has to have its name decorated by {\tt
\_\_declspec(dllexport)}. I've done this for most routines in the AGAMA library by adding

{\tt\#define EXP \_\_declspec(dllexport)} 

\noindent in header files and then placing {\tt EXP} at
appropriate places. In the case of a class, one writes

{\tt class EXP some\_class$\{$..}

\noindent which will cause VS to put into the {\tt.lib} file objects and
methods defined in the class definition. When a method's code is given
outside the class definition, its name needs another {\tt EXP}:

{\tt EXP double some\_class::some\_method(int i,..)$\{$..}

\noindent even though the method was of course declared when the class was defined.
Failure to decorate the name when the code is given generates a `redefinition with
different linkage' error. If the compiler can figure out the need for decoration, why the
hell can't it add it without bothering me?

\noindent The syntax for a {\tt struct} mirrors that of a {\tt class}

{\tt struct EXP some\_struct$\{$..}

\noindent Functions unconnected with a class are handled like methods define outside a
class

{\tt EXP double do\_something(double x,double y)$\{$..}

\noindent The older of my versions of Visual Studio complained when the names
of symbols in internal namespaces were decorated. Since these symbols should
only be referenced from within the library, this should not be a problem.
Anyway don't add {\tt EXP} within internal namespaces.

If you reference a symbol in the library that's not had its name thus
decorated, it will be missing from the {\tt.lib} file and the linker will
report `unresolved symbols'.  I haven't put {\tt EXP}s in the {\tt raga}
files or {\tt py\_wrapper} because I haven't used these bits of code. I have
run all the {\tt test\_xx} programs in the distributed {\tt tests} folder and
checked that the results are comparable to those obtained (on a different
machine) with Gnu.


\bsk
\line{James Binney\hfil May 2021}
\vfill\eject

\centerline{\titlefont Implementing the Python wrappers on Windows}
\msk\noindent
 Access to much of the AGAMA code from Python has been achieved via the
PYBIND11 package. If you want to use our wrappers, your first step should be
to install PYBIND11 by typing at a command prompt

{\tt pip install PYBIND11

}\noindent
Python responds to {\tt import xx} by following the {\tt PYTHONPATH} and
there looking for a folder xx. If it finds this folder, it tries to open {\tt
\_\_init\_\_.py} within it. If it doesn't find the folder, it tries to open
{\tt xx.pyd} in the {\tt PYTHONPATH}. So after installing PYBIND11, create a
directory {\tt agama} in the {\tt PYTHONPATH} and copy into it our file 
{\tt \_\_init\_\_.py}, which contains
\begintt
import os
os.add_dll_directory(``h:/u/c/agama/agama/x64/release'')
from Py_agama import *
\endtt
Edit the bit in quotes to the path to {\tt agama.dll} and
{\tt gsl.dll} on your machine (if they are in separate folders, use two {\tt
os.add\_dll} statements).

Copy our files {Py\_agama.cpp} and {Py\_agama.sln} into a folder and start a
new project in VS by navigating to this folder and clicking on
{Py\_agama.sln}. Now build the solution. This should yield {\tt
Py\_agama.pyd} in the {\tt x64/release} folder. Copy this to the folder
pointed to by PYTHONPATH and you are all set.

\bsk
\line{James Binney \& Tom Wright\hfil June 2025}



\bye 